Transition path sampling based calculation of free energies of enzymatic reactions

### 2022-04-14 tex/manuscript.tex
added a lot more text now

TPS in particular is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
complex for large biological systems.
TPS is based on the Monte Carlo importance sampling in the 
space of trajectories where the collected ensemble of 
reactive pathways are real dynamical trajectories generated without the 
application of any bias forces. Thus the TPS ensemble can be used to obtain 
kinetic information such as the rate of the reaction as well as to elucidate 
the transition state, reaction coordinate and the reaction mechanism. 
TPS has been successfully applied to study problems such as enzyme catalysis, 
protein folding, etc.

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS needs more attention. 
Since the TPS path ensemble only consists of reactive trajectories, the information from 
the discarded non-reactive trajectories which is required for obtaining the 
equilibrium probability distribution of the order parameter is missing. \cite{Dellago09AdvCompSimAppp167}

Methods that can be used to simultaneously calculate
the dynamical as well as equilibrium properties of a system within 
a certain set of asssumptions have garnered interest in recent times. \cite{Suarez14JChemTheoryComput10p2658}
Transition interface sampling (TIS) is a variant of TPS which can be used for 
this purpose. \cite{vanErp03JChemPhys118p7762}

TPS based free energy simulations have been attempted previously by 
Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111,Dellago04MolSimul30p795}

In this manuscript we will focus on the implementation and application of the 
free energy based calculations within the TPS method on reactions catalyzed by two systems:
the human MAT2A enzyme and the plasmodium vivax adenosine deaminase 
enzyme. \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
These enzymes were chosen not only because of their importance in biochemical pathways and
but also because of the availability of experimental results with which comparisons can 
be drawn.

One example of the TPS based free energies calculations on real 
protein systems \cite{Silveira21JPhysChemB125p2018} Human Mat2a enzyme and 
plasmodium vivax adenosine deaminase 
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
Here we briefly describe the equations describing the TPS method for the purpose of
introducing notations as well as for the sake of being complete. 
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is run for a total 
time of $\mathcal{T}$ the number of time slices is given by $L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

We used the crystal structure of human MAT2A 

@@

(7L1A from the protein data bank (PDB)) reported by Niland 

@@

in complex with MT-coformycin (3EWC from the PDB) reported by Ho et 

@@

simulations. Molecular dynamics simulations within classical and hybrid 
quantum mechanics/molecular mechanics (QMMM) approximations were performed 
using the CHARMM program package and using the CHARMM 36 force 
field \cite{Brooks09JComputChem30p1545} for the MM part and the 
approximate PM3 semiempirical method \cite{Repasky02JComputChem23p1601} for the QM part.

@@

We used transition path sampling to generate and
analyze the ensemble of reactive trajectories. As order
parameters that define the reactant and product basins, we
used the donor-hydride and acceptor-hydride distances. The
reactant basin was defined as a donor-hydride distance smaller
than 1.6  and the product basin as an acceptor-hydride
distance smaller than 1.6 . To seed the generation of the path
ensemble, we generated a biased trajectory by applying a
harmonic force between the hydride and the acceptor and
allowed the system to propagate for 250 fs. New trajectories
were generated by the shooting algorithm: we perturbed the
momenta of a randomly chosen slice of a reactive trajectory,
and using the new momenta, we allowed the system to
propagate forward and backward in time for 250 fs. A total of
120 reactive trajectories were generated.
%------------------------------------------------------------------------------ 
\subsection{Plasmodium vivax adenosine deaminase}
%------------------------------------------------------------------------------

%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.

>>>>

TPS in particular is an attractive method since it samples reactive trajectories
in which the rare barrier-crossing events are guaranteed to occur and
it does not require a priori knowledge of the reaction coordinate which can be 
complex for large biological systems.
TPS is based on the Monte Carlo importance sampling in the 
space of trajectories. It uses the shooting and shifting algorithm 
to collect an ensemble of reactive pathways that are real dynamical 
trajectories generated without the application of any bias forces. \cite{dellago02AdvChemPhys123} 
Therefore the TPS ensemble can be used to obtain 
kinetic information such as the rate of the reaction as well as to elucidate 
the transition state, reaction coordinate and the reaction mechanism. 
TPS has been successfully applied to study problems such as enzyme catalysis, 
\cite{Antoniou01JPhysChemB105p5553,Hay12NatChem4p161,Schwartz09NatChemBiol5p551} 
protein folding, \cite{Bolhuis03ProcNatlAcadSci100p12129,Juraszek12ChemPhys396p30} etc.

Calculating equilibrium properties such as the free energy of an enzymatic 
catalysis reaction within TPS is not straighforward. 
The Monte Carlo algorithm used in TPS has an acceptance criterion that 
is only satisfied if the trajectory begins in the reactant basin and ends in 
the product basin or vice versa, hence trajectories that are localized in the 
reactant or product basins are rejected. The localized trajectories contribute
to the probability density distribution function ($P(\lambda)$) which is the 
probability to find the system at a particular value of the order parameter $\lambda$.

Since the TPS path ensemble only consists of reactive trajectories, the information from 
the discarded non-reactive trajectories which is required for obtaining the 
equilibrium probability distribution of the order parameters is 
missing. \cite{Dellago09AdvCompSimAppp167} Radhakrishnan et al. developed 
`BOLAS' which combines the shooting and shifting algorithms of 
TPS with a window based sampling technique 
in the spirit of umbrella sampling to calculate equilibrium free energy 
differences. \cite{Radhakrishnan04JChemPhys121p2436} 
Baron peters developed a variant of this scheme called the 
equilibrium path sampling (EPS) where they use aimless shooting 
to create the ensemble within each window. \cite{Peters08JAmChemSoc130p17342,Silveira21JPhysChemB125p2018} 
Apart from these schemes a recent work by 
Brotzakis et al., \cite{Brotzakis19JChemPhys151p174111,Dellago04MolSimul30p795}
has attempted to look at the TPS based free energy calculations for 
reactions that are diffusive in nature where 
Transition interface sampling (TIS) \cite{vanErp03JChemPhys118p7762}
could work well.

In this manuscript we will focus on the implementation and application of the 
free energy based calculations within the TPS method on reactions catalyzed by two systems:
the human MAT2A enzyme and the plasmodium vivax adenosine deaminase 
enzyme. \cite{Luo07JAmChemSoc129p8008,Ho09Biochemistry48p9618}
These enzymes were chosen not only because of their importance in biochemical pathways and
but also because of the availability of experimental results with which comparisons can 
be drawn. 
%------------------------------------------------------------------------------ 
\section{Methods}
%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------ 
Here we briefly describe the equations describing the TPS method for the purpose of
introducing notations as well as for the sake of being complete. 
Consider the molecular dynamics simulation of a molecular system consisting 
of $N$ atoms. Starting from the initial positions and momenta of each atom at time $t_0$,
the time evolution is carried out by finding the position ($\textbf{q}$) and 
momenta ($\textbf{p}$) of each atom at regular intervals of time $t_i = t_0 + i\Delta t$  
($i = 0,1,2,\ldots$). At each of the time slices, the positions and 
momenta can be collectively represented as
the set $z = \{\textbf{q},\textbf{p}\}$. If the MD simulation is 
run for a total 
time of $\mathcal{T}$ the number of time slices is given by 
$L = \mathcal{T}/\Delta t +1$ and for this sequence of 
times the state of the system can be represented as 
\begin{equation}
z(\mathcal{T}) = \{z_0, z_{\Delta t}, z_{2\Delta t},\ldots,z_{\mathcal{T}}\}
\end{equation}
The probability to obtain a particular sequence of states is determined by the initial 
conditions and the type of dynamics used for the time evolution. If the time evolution is 
Markovian, the probability to go from $z_{i\Delta t}$ to $z_{(i+1)\Delta t}$ depends only 
on $z_{i\Delta t}$ and not on conditions before the time $i\Delta t$ the total path probablility can 
be expressed as the product of individual probabilities $p(i\rightarrow j)$ as 
\begin{equation}
\mathcal{P}[z(\mathcal{T})] = \rho(z_0)\Pi_{i=0}^{\mathcal{T}/\Delta t-1} p(z_{i\Delta t}\rightarrow z_{(i+1)\Delta t}),
\end{equation}
where $\rho(z_0)$ is the equilibrium distribution of the initial conditions and for a canonical ensemble this can be 
expressed as 
\begin{equation}
\rho(z_0) = \exp(-\beta H(z_0))/Z
\end{equation}
where 
\begin{equation}
Z = \int dz \exp(-\beta H(z)) 
\end{equation}
is the canonical partition function.

@@

We used the crystal structure of the human MAT2A enzyme 

@@

(7L1A from the protein data bank) reported by Niland 

@@

in complex with MT-coformycin (3EWC from the protein data bank) reported by Ho et 

@@

simulations. For the human MAT2A enzyme the substrates were modified to 
S-Adenosyl methionine (SAM) and adenosyl triphosphate (ATP) whereas for the \textit{pv}ADA
enzyme the substrates were modified to $2^{'}-$deoxyadenosine and hydroxide ion.

MD simulations within classical and hybrid 
quantum mechanics/molecular mechanics (QMMM) approximations were carried out 
using the CHARMM program package. \cite{Brooks83JComputChem4p187,Brooks09JComputChem30p1545} 
For the MM calculations the CHARMM36 force field \cite{Brooks09JComputChem30p1545} 
was used and for the QM calculations the approximate PM3 semiempirical 
method \cite{Repasky02JComputChem23p1601} was used. Both the
systems were solvated with water molecules using the TIP3P model
in a nanodroplet, whose volume was set to be $15\;\AA$ away from
the surface of the protein, and then neutralized with potassium
ions.

@@

We used transition path sampling to generate and
analyze the ensemble of reactive trajectories. As order
parameters that define the reactant and product basins, we
used the donor-hydride and acceptor-hydride distances. The
reactant basin was defined as a donor-hydride distance smaller
than 1.6  and the product basin as an acceptor-hydride
distance smaller than 1.6 . To seed the generation of the path
ensemble, we generated a biased trajectory by applying a
harmonic force between the hydride and the acceptor and
allowed the system to propagate for 250 fs. New trajectories
were generated by the shooting algorithm: we perturbed the
momenta of a randomly chosen slice of a reactive trajectory,
and using the new momenta, we allowed the system to
propagate forward and backward in time for 250 fs. A total of
120 reactive trajectories were generated.
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.
